\large How resource abundance and stochasticity affect organisms' space-use requirements

### 2023-11-23 writing/manuscript.Rmd
prepared manuscript and appendices for submission

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

<!-- place authors' emails on a new page -->

\noindent \textbf{Authors' emails:}

\noindent Stefano Mezzini: stefano.mezzini@ubc.ca

\noindent Chris H. Fleming: christen.fleming@ucf.edu

\noindent E. Patrícia Medici: medici@ipe.org.br

\noindent Michael J. Noonan: michael.noonan@ubc.ca

<!-- place table of contents on a new page -->
\newpage

\singlespacing

\tableofcontents

@@

\noindent The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike. In this study, we first review the literature of studies that estimate organisms' space-use requirements as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

\vspace*{\fill}

\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`.

@@

\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the **resources** the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are typically reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements. In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

Here, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource *abundance* on organisms' space use while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' space use while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix 1.

@@

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean). (a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color). (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see fig. \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (fig. \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.

@@

Like the scenario illustrated in fig. \@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time. We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (fig. \@ref(fig:multipanel-changing-var)). When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex. Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$. Biologically, this is important because some organisms may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event. An organism's perception of changes in $R$ will depend strongly on the organism's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013]. The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires organisms have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$. Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean). (a.) Although resources ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

@@

Each of the 200 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point). Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution. The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig. \@ref(fig:5-5-sims)). Note that $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from. Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ with no correlation to the following track. Finally, each set of 200 truncated tracks was modeled using an OUF model and Autocorrelated Kernel Density Estimate to estimate the 95% utilization distribution. Additional information is provided in Appendix 2.

@@

Fig. \@ref(fig:5-5-sims) shows how the simulated space-use requirements changed in response to $\mu(t)$ and $\sigma^2(t)$. The top row (constant $\var(R)$) shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig. \@ref(fig:multipanel-constant-var)). As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the space-use requirements decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism. As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].

@@

Although the temporal trends in fig. \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time. Both $\e(R)$ and $\var(R)$ affect space-use requirements precisely as we hypothesized (fig. \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero). We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of fig. \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors. Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$. Secondly, the organism's movement was simulated to be range resident within an homogeneous environment. Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, organisms may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood. Thus, we hypothesize that our simulations may be underestimating organisms' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated organisms cannot select for areas of higher $\e(R)$ or lower $\var(R)$. Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$. While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.

@@

\noindent In this section, we show how this framework can be applied to better understand the movement ecology of free-ranging organisms via remote sensing data such as NDVI [@pettorelli_normalized_2011]. To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado. Appendix 3 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.

@@

\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the **linear transformation**

@@

\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data]. <!--Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

@@

# Acknowledgements

@@

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
# author:
#   - name: Stefano Mezzini
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
#     correspondence: "yes"
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
# biblio-style: 'amnatnat'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

\bibliographystyle{amnatnat}

@@

```{r, echo=FALSE}
# needs to be <= 200 words for AmNat
abstract <- "The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike. In this study, we leverage the available literature to provide a unifying framework and hypotheses for the effect of mean and variance in resources on organisms' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
```

<!-- elements of the title page -->

\noindent \textbf{Article type}: Major article

\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`

\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`

\noindent \textbf{Figures}: 5

\noindent \textbf{Tables}: 0

\noindent \textbf{References}: 141 <!-- counted by using nature csl -->

\noindent \textbf{Appendices}: 3

\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`

<!-- Authors' emails:
Stefano Mezzini: stefano.mezzini@ubc.ca
Chris H. Fleming: christen.fleming@ucf.edu
E. Patrícia Medici: medici@ipe.org.br
Michael J. Noonan: michael.noonan@ubc.ca
-->

@@

\noindent `r abstract`

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are typically reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements. In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

Here, we refer to a location's average amount of resources as resource abundance, while we use the phrase resource stochasticity to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource abundance on organisms' space use while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource stochasticity on organisms' space use while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix A.

@@

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see fig. A1). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (fig. A2). Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict where $R$ is high if $\e(R)$ does not vary predictably over space.

@@

Like the scenario illustrated in fig. A1, we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time. We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (fig. A2). When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex. Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$. Biologically, this is important because some organisms may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event. An organism's perception of changes in $R$ will depend strongly on the organism's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013]. The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires organisms have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$. Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.

@@

Each of the 200 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point). Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution. The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig. \@ref(fig:5-5-sims)). Note that $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from. Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ with no correlation to the following track. Finally, each set of 200 truncated tracks was modeled using an OUF model and Autocorrelated Kernel Density Estimate to estimate the 95% utilization distribution. Additional information is provided in Appendix B.

@@

Fig. \@ref(fig:5-5-sims) shows how the simulated space-use requirements changed in response to $\mu(t)$ and $\sigma^2(t)$. The top row (constant $\var(R)$) shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig. A1). As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the space-use requirements decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism. As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].

@@

Although the temporal trends in fig. \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time. Both $\e(R)$ and $\var(R)$ affect space-use requirements precisely as we hypothesized (fig. \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero). We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of fig. \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors. Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases sublinearly with $\var(R)$. Secondly, the organism's movement was simulated to be range resident within an homogeneous environment. Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, organisms may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood. Thus, we hypothesize that our simulations may be underestimating organisms' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated organisms cannot select for areas of higher $\e(R)$ or lower $\var(R)$. Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$. While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.

@@

\noindent In this section, we show how this framework can be applied to better understand the movement ecology of free-ranging organisms via remote sensing data such as NDVI [@pettorelli_normalized_2011]. To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado. Appendix C contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.

@@

\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the linear transformation

@@

\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data]. <!--Note that only linear transformations (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$. We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is

@@

<!--
# Acknowledgements

@@

# Author contribution statement
After acceptance, all multi-author papers must provide an Author Contribution Statement in the final files. It appears after the Acknowledgments. The goal of the Author Contribution statement is to clearly assign responsibility for parts of the paper. This statement should be carefully written to indicate the role that every author played in generating the paper. However, it is essential that authors do not get credited for steps in which they were not deeply involved. Sample roles are provided below but be sure to assemble a Contribution Statement that fits the specifics of your paper:
